### 2018-02-20 paper/sections/1_introduction.tex
paper: address reviewer feedback
Signed-off-by: Michael Sevilla <mikesevilla3@gmail.com>

\begin{abstract}

Our analysis of the key-value activity generated by the ParSplice molecular
dynamics simulation demonstrates the need for more complex cache management
strategies. Baseline measurements show clear key access patterns and hot spots
that offer significant opportunity for optimization. We use the data management
language and policy engine from the Mantle system to dynamically explore a
variety of techniques, ranging from basic algorithms and heuristics to
statistical models, calculus, and machine learning. While Mantle was originally
designed for distributed file systems, we show how the collection of
abstractions effectively decomposes the problem into manageable policies for a
different application and storage system.  Our exploration of this space results in a
dynamically sized cache policy that does not sacrifice any performance while
using 32-66\% less memory than the default ParSplice configuration.

\end{abstract}

@@

The ParSplice molecular dynamics simulation is representative of an important
class of HPC applications with similar working set behaviors that extensively
use software-based caches. It uses a hierarchy of caches and a single
persistent key-value store to store both observed minima across a molecule's
equation of motion (EOM) and the hundreds or thousands of partial trajectories
calculated each second during a parallel job.  This workload is pervasive
across simulations that (1) rely on a mesh-based decomposition of a physical
region and (2) result in millions or billions of mesh cells, where each cell
contains materials, pressures, temperatures and other characteristics that are
required to accurately simulate phenomena of interest.  The fine-grained data
annotation capabilities provided by key-value storage is a natural match for
these types of scientific simulations.  Unfortunately, simulations of this size
saturate the capacity and bandwidth capabilities of a single node so we need
more effective data management techniques.

\begin{figure}[t]
\noindent\includegraphics[width=0.5\textwidth]{figures/cache-management.png}\\
\caption{Using our data management language and policy engine, we design a
dynamically sized caching policy (thick line) for ParSplice.  Compared to
existing configurations (thin lines with \(\times\)'s), our solution saves the most
memory without sacrificing performance and works for a variety of inputs.
\label{fig:cache-management}}
\end{figure}

The biggest challenge for ParSplice is properly sizing the caches in the
storage hierarchy.  The memory usage for a single cache that stores molecule
coordinates is shown in Figure~\ref{fig:cache-management}, where the thin solid
lines marked with \(\times\)'s are the existing configurations in ParSplice.
The default configuration uses an unlimited sized cache, shown by the ``No
Cache Management" line, but using this much memory for one cache is
unacceptable for HPC environments, where a common goal is to keep memory for
such data structures below 3\%\footnote{Anecdotally, this threshold works well
for HPC applications.  For reference, a 1GB cache for a distributed file system
is too large in LANL deployments.}. Furthermore, ParSplice deploys a cache per
300 worker processes, so large simulations need more caches and will use even
more memory.  Users can configure ParSplice to evict data when the cache
reaches a threshold but this solution requires tuning and parameter sweeps; the
``Cache (too small)" curve in Figure~\ref{fig:cache-management} shows how a
poorly configured cache can save memory but at the cost of performance, which
is shown by the text annotation to the right.  Even worse, this threshold
changes with different initial configurations and cluster setups so tuning
needs to be done for all system permutations.  Our dynamically sized cache,
shown by the thick line in Figure~\ref{fig:cache-management}, detects key
access patterns and re-sizes the cache accordingly.  Without tuning or
parameter sweeps, our solution saves more memory than a hand-tuned cache
without any performance degradation, works for a variety of initial conditions,
and could generalize to similar applications.

@@

This last contribution is explored in Sections~\S\ref{sec:arch-specific}
and~\S\ref{sec:dom-specific}, where we try a range of policies from different
disciplines; but more importantly, in Section~\S\ref{sec:scope}, we conclude
that the collection of policies we design for cache management in ParSplice are
very similar to the policies used to load balance metadata in the Ceph file
system (CephFS~\cite{weil:osdi2006-ceph}) suggesting that there is potential
for automatically adapting and generating policies dynamically.

>>>>

@@

ParSplice molecular dynamics simulations are representative of an important
class of HPC applications with similar working set behaviors that extensively
use software-based caches. ParSplice uses a hierarchy of caches and a single
persistent key-value store to store the atomic coordinates corresponding
to local energy minima (referred to simply as minima) encountered
by a large number of independent dynamical trajectories.
This workload is also pervasive
across simulations that (1) rely on a mesh-based decomposition of a physical
region and (2) result in millions or billions of mesh cells, where each cell
contains materials, pressures, temperatures and other characteristics that are
required to accurately simulate phenomena of interest.  The fine-grained data
annotation capabilities provided by key-value storage is a natural match for
these types of scientific simulations.  Unfortunately, simulations of this size
saturate the capacity and bandwidth capabilities of a single node so we need
more effective data management techniques.

\begin{figure}[t]
\noindent\includegraphics[width=0.5\textwidth]{figures/cache-management.png}\\
\caption{Using our data management language and policy engine, we design a
dynamically sized caching policy (thick line) for ParSplice.  Compared to
existing configurations (thin lines with \(\times\)'s), our solution saves the most
memory without sacrificing performance and works for a variety of inputs.
Solutions labeled ``no performance degradation" are in comparison to the
baseline of using a cache of unlimited size (``No Cache Management").
\label{fig:cache-management}}
\end{figure}

A challenge for ParSplice is properly sizing the caches in the
storage hierarchy.  The memory usage for a single cache that stores atomic
coordinates is shown in Figure~\ref{fig:cache-management}, where the thin solid
lines marked with \(\times\)'s are the existing configurations in ParSplice.
The default configuration uses an unlimited sized cache, shown by the ``No
Cache Management" line, but using this much memory for one cache is
unacceptable for HPC environments, where a common goal is to keep memory for
such data structures below 3\%\footnote{Anecdotally, this threshold works well
for HPC applications.  For reference, a 1GB cache for a distributed file system
is too large in LANL deployments.}. Furthermore, ParSplice deploys a cache per
300 worker processes, so large simulations need more caches and will use even
more memory.  Users can configure ParSplice to evict data when the cache
reaches a threshold but this solution requires tuning and parameter sweeps; the
``Cache (too small)" curve in Figure~\ref{fig:cache-management} shows how a
poorly configured cache can save memory but at the cost of performance, which
is shown by the text annotation to the right.  Even worse, this threshold
changes with different initial configurations and cluster setups so tuning
needs to be done for all system permutations.  Our dynamically sized cache,
shown by the thick line in Figure~\ref{fig:cache-management}, detects key
access patterns and re-sizes the cache accordingly.  Without tuning or
parameter sweeps, our solution saves more memory than a hand-tuned cache
without any performance degradation, works for a variety of initial conditions,
and could generalize to similar applications. Triggering key eviction at a
certain memory pressure ({\it e.g.} 3\%) requires {\it a priori} system
knowledge while our approach saves the most memory because we model the
behavior of keyspace accesses.

@@

This last contribution is explored in Sections~\S\ref{sec:arch-specific}
and~\S\ref{sec:dom-specific}, where we try a range of policies from different
disciplines. We analyze why our early policies fail for our HPC hardware and
software and show how we iterated to the final solution.  More importantly, in
Section~\S\ref{sec:scope}, we conclude that the collection of policies we
design for cache management in ParSplice are very similar to the policies used
to load balance metadata in the Ceph file system
(CephFS~\cite{weil:osdi2006-ceph}) suggesting that there is potential for
automatically adapting and generating policies dynamically. This paper does not
thoroughly study cache management policies for file systems, but our findings
for this specific application and related storage systems work suggests that
this approach is a strong avenue for future work.

### 2018-02-20 paper/sections/2_parsplice-keyspace-analysis.tex
paper: address reviewer feedback
Signed-off-by: Michael Sevilla <mikesevilla3@gmail.com>

ParSplice~\cite{perez:jctc20150parsplice} is an accelerated molecular dynamics
(MD) simulation package developed at LANL. It is part of the Exascale Computing
Project\footnote{http://www.exascale.org/bdec/} and is important to LANL's
Materials for the Future initiative.

@@

The computation can be parallelized by adding more workers or by adding tasks
to parallelize individual workers.  The workers are stateless and read initial
coordinates from the data store each time they begin generating segments. Since
worker tasks do not maintain their own history, they can end up reading the
same coordinates repeatedly. To mitigate the consequences of these repeated
reads, ParSplice provisions a hierarchy of caches that sit in front of a single
persistent database.  Values are written to each tier and reads traverse up the
hierarchy until they find the data.

We use ParSplice to simulate the evolution of metallic nanoparticles that grow
from the vapor phase.  This simulation stresses the storage hierarchy more than
other input decks because it uses a cheap potential, has a small number of
atoms, and operates in a complex energy landscape with many accessible states.
As the run progresses, the energy landscape of the system becomes more complex
and more states are visited.  Two domain factors control the number of entries
in the data store: the growth rate and the temperature. The growth rate
controls how quickly new atoms are added to the nanoparticle: fast growth rates
lead to non-equilibrium conditions, and hence increase the number of states
that can be visited.  However, as the particle grows, the simulation slows down
because the calculations become more expensive, limiting the rate at which new
states are visited.  On the other hand, the temperature controls how easily a
trajectory can jump from state to state; higher temperatures lead to more
frequent transitions but temperatures that are too high result in meaningless
simulations because trajectories have so much energy that they are equally
likely to visit any random state.

@@

\begin{figure}[t]
  \noindent\includegraphics[width=0.5\textwidth]{figures/methodology-keyspace.png}\\
  \caption{The keyspace is small but must satisfy many reads as workers
  calculate segments. Memory usage scales linearly, so it is likely that we will need
  more than one node to manage segment coordinates when we scale the system or jobs up.
  \label{fig:methodology-keyspace}}
\end{figure}

@@

\subsubsection{\underline{Initial conditions influence key activity}}
Figure~\ref{fig:motivation-regimes} shows how ParSplice tasks read key-value
pairs from the worker's cache for two different initial conditions of
\(\Delta\), which is the rate that new atoms enter the simulation.  The line is
the read request rate (\(y1\) axis) and the dots along the bottom are the
number of unique keys accessed (\(y2\) axis).  The access patterns for
different growth rates have temporal locality, as the reads per second for
\(\Delta_2\) look like the reads per second for \(\Delta_1\) stretched out
along the time axis.  The \(\Delta_1\) growth rate adds atoms every 100K
microseconds while the \(\Delta_2\) growth rate adds atoms every 1 million
microseconds. So \(\Delta_2\) has a smaller growth rate resulting in hotter
keys and a smaller keyspace.  Values smaller than \(\Delta_2\)'s growth rate or
a temperature of 400 degrees result in very little database activity because
state transitions take too long. Similarly, values larger than \(\Delta_1\)'s
growth rate or a temperature of 4000 degrees result in an equally meaningless
simulation as transitions are unrealistic.

This figure demonstrates that small changes to \(\Delta\) can have a strong
effect on the timing and frequency with which new EOM minima are discovered and
referenced.  Trends also exist for temperature and number of workers but are
omitted here for space.  This finding suggests that we need a flexible policy
language and engine to explore these trade-offs.

>>>>

ParSplice~\cite{perez:jctc20150parsplice} is an accelerated molecular dynamics
(MD) simulation package developed at LANL. It is part of the Exascale Computing
Project\footnote{https://www.exascaleproject.org/} and is important to LANL's
Materials for the Future initiative.

@@

The computation can be parallelized by adding more workers or by adding tasks
to parallelize individual workers.  The workers are stateless and read initial
coordinates from the data store each time they begin generating segments. Since
worker tasks do not maintain their own history, they can end up reading the
same coordinates repeatedly. To mitigate the consequences of these repeated
reads, ParSplice provisions a hierarchy of caches that sit in front of a single
persistent database.  Values corresponding to new keys are written to each tier
and reads traverse up the hierarchy until they find the data.

We use ParSplice to simulate the evolution of metallic nanoparticles that grow
from the vapor phase.  This simulation stresses the storage hierarchy more than
other input decks because it uses a cheap potential, has a small number of
atoms, and operates in a complex energy landscape with many accessible states.
As the run progresses, the energy landscape of the system becomes more complex
and more states are visited.  Two domain factors control the number of entries
in the data store: the growth rate and the temperature. The growth rate
controls how quickly new atoms are added to the nanoparticle: fast growth rates
lead to non-equilibrium conditions, and hence increase the number of states
that can be visited.  However, as the particle grows, the simulation slows down
because the calculations become more expensive, limiting the rate at which new
states are visited.  On the other hand, the temperature controls how easily a
trajectory can jump from state to state; higher temperatures lead to more
frequent transitions but temperatures that are too high lead to the melting of the
nanoparticle and hence to a complete change in the physics of the system.

@@

\begin{figure}[t]
  \noindent\includegraphics[width=0.5\textwidth]{figures/methodology-keyspace.png}\\
  \caption{The keyspace is small but must satisfy many reads as workers
  calculate segments. Memory usage scales linearly, so it is likely that we will
  need more than one node to manage segment coordinates when we scale the system
  or jobs up. \(\Delta_1\) is the growth rate, an application specific parameter
  described in Section~\S\ref{sec:delta}.
  \label{fig:methodology-keyspace}}
\end{figure}

@@

\subsubsection{\underline{Initial conditions influence key activity}}
\label{sec:delta}
Figure~\ref{fig:motivation-regimes} shows how ParSplice tasks read key-value
pairs from the worker's cache for two different initial conditions of
\(\Delta\), which is the rate that new atoms enter the simulation.  The line is
the read request rate (\(y1\) axis) and the dots along the bottom are the
number of unique keys accessed (\(y2\) axis).  The access patterns for
different growth rates have temporal locality, as the reads per second for
\(\Delta_2\) look like the reads per second for \(\Delta_1\) stretched out
along the time axis.  The \(\Delta_1\) growth rate adds atoms every 100K
picoseconds while the \(\Delta_2\) growth rate adds atoms every 1 million
picoseconds. So \(\Delta_2\) has a smaller growth rate resulting in hotter
keys and a smaller keyspace.  Values smaller than \(\Delta_2\)'s growth rate or
a temperature of 400 degrees result in very little database activity because
state transitions take very long. Similarly, growth rate values larger than \(\Delta_1\)
lead to far out-of-equilibrium growth, while temperatures in excess of 800 degrees
result in the melting of the nanoparticle, which is not relevant in this application.

This figure demonstrates that small changes to \(\Delta\) can have a strong
effect on the timing and frequency with which new minima are discovered and
referenced.  Trends also exist for temperature and number of workers but are
omitted here for space.  This finding suggests that we need a flexible policy
language and engine to explore these trade-offs.

### 2018-02-20 paper/sections/4_dynamic-policies.tex
paper: address reviewer feedback
Signed-off-by: Michael Sevilla <mikesevilla3@gmail.com>

% technical details
We implement a basic LRU cache using a ``when" policy of:

\begin{figure}[h]
\footnotesize
\begin{minted}{lua}
server[whoami]['cachesize']>n
\end{minted}
\end{figure}

and a ``how much" policy of:

\begin{figure}[h]
\footnotesize
\begin{minted}{lua}
servers[whoami]['cachesize']-n
\end{minted}
\end{figure}

The results for different cache sizes for a growth rate of \(\Delta_1\) over a
2.5 hour run across 256 workers is shown in
Figure~\ref{fig:methodology-tradeoff}.  ``Baseline" is the performance of
unmodified ParSplice  measured in trajectory duration (\(y1\) axis) and
utilization is measured with memory footprint of just the cache (\(y2\) axis).
The middle graph labeled ``Fixed Cache Size" shares the \(y\) axes and shows
the trade-off of using a basic LRU-style cache of different sizes, where the
penalty for a cache miss is retrieving the data from the persistent database.
The error bars are the standard deviation of 3 runs.  Although the keyspace
grows to 150K, a 100K key cache achieves 99\% of the performance. Decreasing
the cache degrades performance and predictability.

@@

\caption{Key activity for a 4 hour run shows groups of accesses to the same
subset of keys. Detecting these access patterns leads to a more accurate cache
management strategy, which is discussed in Section~\S\ref{sec:regime-detection}
and the results are in Figure~\ref{fig:dscache-vs-none}.
\label{fig:keyspace-zoomed}}
    \end{subfigure}%
    ~ 
    \begin{subfigure}[t]{0.48\textwidth}
        \centering
        \includegraphics[width=1\textwidth]{figures/dscache-vs-none.png}\\

	\caption{The performance/utilization for the dynamically sized cache (DSCache)
policy. With negligible performance degradation, DSCache adjusts to different initial configurations
(\(\Delta\)'s) and saves 3\(\times\) as much memory in the best case.
\label{fig:dscache-vs-none}}

    \end{subfigure}%
    \caption{Different cache management policies tested over the Mantle policy engine.}
\end{figure*}

@@

Figure~\ref{fig:keyspace-zoomed} shows which keys (\(y\) axis) are accessed by
tasks over time (\(x\) axis). The groups of accesses to a subset of keys occurs
because molecules are stuck in deep trajectories. Recall that the cache stores
the molecules' EOM minima, which is the smallest effective energy that a
molecule observes during its trajectory. So molecules stuck in deep
trajectories explore the same minima until they can escape to a new set of
states. This exploration of the same set of states is called a superbasin.  In
Figure~\ref{fig:keyspace-zoomed}, superbasins are never re-visited because the
simulation only adds molecules; we can never reach a state with less molecules.
This is why keys are never re-accessed.

>>>>

% technical details
We implement a basic LRU cache using a ``when" policy of:
\texttt{server[whoami]['cachesize']>n} and a ``how much" policy of
\texttt{servers[whoami]['cachesize']-n}.  The results for different cache sizes
for a growth rate of \(\Delta_1\) over a 2.5 hour run across 256 workers is
shown in Figure~\ref{fig:methodology-tradeoff}.  ``Baseline" is the performance
of unmodified ParSplice  measured in trajectory duration (\(y1\) axis) and
utilization is measured with memory footprint of just the cache (\(y2\) axis).
The middle graph labeled ``Fixed Cache Size" shares the \(y\) axes and shows
the trade-off of using a basic LRU-style cache of different sizes, where the
penalty for a cache miss is retrieving the data from the persistent database.
The error bars are the standard deviation of 3 runs.  Although the keyspace
grows to 150K, a 100K key cache achieves 99\% of the performance. Decreasing
the cache degrades performance and predictability.

@@

\caption{Key activity for a 4 hour run shows groups of accesses to the same
subset of keys. Detecting these access patterns leads to a more accurate cache
management strategy, which is discussed in Section~\S\ref{sec:regime-detection}
and the results are in Figure~\ref{fig:dscache-vs-none}.
\label{fig:keyspace-zoomed}}
    \end{subfigure}%
    ~ 
    \begin{subfigure}[t]{0.48\textwidth}
        \centering
        \includegraphics[width=1\textwidth]{figures/dscache-vs-none.png}\\
	\caption{The performance/utilization for the dynamically sized cache (DSCache)
policy. With negligible performance degradation, DSCache adjusts to different initial configurations
(\(\Delta\)s) and saves 3\(\times\) as much memory in the best case.
\label{fig:dscache-vs-none}}
    \end{subfigure}%
    \caption{Different cache management policies tested over the Mantle policy engine.}
\end{figure*}

@@

Figure~\ref{fig:keyspace-zoomed} shows which keys (\(y\) axis) are accessed by
tasks over time (\(x\) axis). The groups of accesses to a subset of keys occurs
because the system is stuck in so-called superbasins, {\it i.e.} coarse regions of space from
which it is difficult to escape, but within which it is easy to move. Systems
stuck in superbasins will explore the same set of minima for a
long time, leading to the same keys being accessed repeatedly. In
Figure~\ref{fig:keyspace-zoomed}, superbasins are typically not re-visited because the
simulation only adds atoms; we can never revisit a superbasin that contains less atoms
than the simulation currently contains.
This is why keys are never re-accessed after a given amount of time.

### 2018-02-20 paper/sections/6_conclusion.tex
paper: address reviewer feedback
Signed-off-by: Michael Sevilla <mikesevilla3@gmail.com>

% ml and autotuning
Auto-tuning is a well-known technique used in
HPC~\cite{behzad:sc2013-autotuning, behzad:techreport2014-io-autotuning}, big
data systems systems~\cite{herodotou_starfish_2011}, and
databases~\cite{schnaitter_index_2009}.  Like our work, these systems focus on
the physical design of the storage ({\it e.g.} cache size) but since we focused
on a relatively small set of parameters (cache size, migration thresholds), we
did not need anything as sophisticated as the genetic algorithm used
in~\cite{behzad:sc2013-autotuning}.  We cannot drop these techniques into
ParSplice because the magnitude and speed of the workload hotspots/flash crowds
makes existing approaches less applicable.

>>>>

% ml and autotuning
Auto-tuning is a well-known technique used in
HPC~\cite{behzad:sc2013-autotuning}, big data systems
systems~\cite{herodotou_starfish_2011}, and
databases~\cite{schnaitter_index_2009}.  Like our work, these systems focus on
the physical design of the storage ({\it e.g.} cache size) but since we focused
on a relatively small set of parameters (cache size, migration thresholds), we
did not need anything as sophisticated as the genetic algorithm used
in~\cite{behzad:sc2013-autotuning}.  
%We cannot drop these techniques into
%ParSplice because the magnitude and speed of the workload hotspots/flash crowds
%makes existing approaches less applicable.
